---
title: "Statistical-Modelling-Project"
author: "Stephen Fekete"
date: "2025-05-29"
output: word_document
---

## Project Summary

This project explores the relationship between the Southern Oscillation Index (SOI) and seasonal rainfall across Australia. 
Using R, we perform multiple linear regression, diagnostic analysis, and model refinement to improve predictive accuracy.


## Load and Inspect Data

```{r}
# Load required libraries
library(tidyverse)
library(broom)

# Read in data
soi_df <- read_csv("seasonal_soi_data.csv")
rain_df <- read_csv("total_seasonal_rainfall.csv")

# Merge datasets
df <- inner_join(soi_df, rain_df, by = c("Year", "Season"))
head(df)
```

This section loads the required datasets, merges them on common keys, and previews the data for further analysis.

## Exploratory Data Analysis

```{r}
# Summary statistics
summary(df)

# Scatter plot by season with null and linear model overlays
ggplot(total_seasonal_rainfall, aes(x = SeasonalSOI, y = total_seas_prcp)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", formula = y ~ 1, se = FALSE, colour = "red", linetype = "dashed") +  # Null model
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "blue") +  # Linear model
  facet_wrap(~ Season) +
  labs(
    title = "Relationship between Seasonal SOI and Rainfall by Season",
    x = "Seasonal SOI",
    y = "Total Seasonal Rainfall (mm)"
  ) +
  theme_bw()
```

This plot compares the relationship between SOI and total rainfall across seasons. The dashed red line represents the null model (mean rainfall), while the solid blue line represents the fitted linear model. A stronger positive relationship is evident in summer, while other seasons show minimal trends.

## Linear Regression Model

```{r}
# Fit model
model <- lm(total_seas_prcp ~ SeasonalSOI, data = df)

# Model summary
summary(model)
```
This section fits a linear regression model to predict total seasonal rainfall using SOI as the predictor and summarizes the model's output.

## Residual Diagnostics

```{r}
# Plot residuals
par(mfrow = c(2, 2))
plot(model)

# Check for normality
qqnorm(resid(model), main = "QQ Plot of Residuals")
qqline(resid(model))
```
Here we validate model assumptions using residual plots and a Q-Q plot to assess normality of residuals.

## Model Refinement

```{r}
# Polynomial model
model_poly <- lm(total_seas_prcp ~ poly(SeasonalSOI, 2), data = df)
summary(model_poly)
```

This section explores a refined polynomial model to capture any potential nonlinear relationships between SOI and rainfall.

## Conclusion

The analysis shows a moderate relationship between SOI and seasonal rainfall. Model diagnostics suggest assumptions are reasonably met, and a polynomial fit may slightly improve accuracy.
